Comparison of Terrain Following and Cut Cell Grids with a Non-Hydrostatic Model

### 2015-02-10 fig-gravityWaves-theta.tex
add thetaDiff[Zoom] plots

\documentclass{article}
\usepackage[paperwidth=5.4in,paperheight=4in,margin=0in]{geometry}
\usepackage{mathptmx}%{times}
\usepackage{graphicx}
\usepackage{siunitx}

\setlength{\parindent}{0pt}
\begin{document}
\includegraphics[height=2in,angle=270]{../gravityWaves-btf/18000/thetaDiff.eps}

\includegraphics[height=2in,angle=270]{../gravityWaves-snapCol/18000/thetaDiff.eps}
\end{document}

>>>>

\documentclass{article}
\usepackage[paperwidth=6.4in,paperheight=4.1in,margin=0in]{geometry}
\usepackage{mathptmx}%{times}
\usepackage{graphicx}
\usepackage{siunitx}

\setlength{\parindent}{0pt}
\begin{document}
\includegraphics[height=2.2in,angle=270]{../gravityWaves-btf/18000/thetaDiff.eps}
\hspace{0.1in}
\includegraphics[height=4in,angle=270]{../gravityWaves-btf/18000/thetaDiffZoom.eps}
\vspace*{0.1in}

\includegraphics[height=2.2in,angle=270]{../gravityWaves-snapCol/18000/thetaDiff.eps}
\hspace{0.1in}
\includegraphics[height=4in,angle=270]{../gravityWaves-snapCol/18000/thetaDiffZoom.eps}
\vspace*{0.1in}

\includegraphics[height=2.2in,angle=270]{../gravityWaves-diagnostics/thetaDiffLegend.eps}
\hspace{0.2in}
\includegraphics[height=3.9in,angle=270]{../gravityWaves-diagnostics/thetaDiffZoomLegend.eps}
\end{document}

### 2015-02-10 msc-publication.tex
plot thetaDiff for gravityWaves
add thetaDiff[Zoom] plots

\section{Introduction}
Representing orography in numerical weather prediction systems is necessary to model downslope winds and local precipitation patterns.  Terrain also affects global circulation, giving rise to planetary and gravity waves.

There are two main approaches to represent orography on a grid: terrain following layers, and cut cells.  With terrain following (TF) layers the terrain's influence decays with height so that the bottommost layers follow the underlying surface closely while the uppermost layers are flat.  TF layers are usually implemented on a rectangular computational grid using transformed coordinates.

\citet{galchen-somerville1975} proposed a basic terrain following (BTF) coordinate defined as 
\begin{equation}
	z = \left( H - h \right) \left( z^\star / H \right) + h
\end{equation}
where, in two dimensions, \(z(x, z^\star)\) is the height of the coordinate surface at level \(z^\star\), \(H\) is the height of the domain, and \(h(x)\) is the height of the terrain surface.  Using this coordinate, the terrain's influence decays linearly with height but disappears only at the top of the domain.

It is well-known that TF coordinates perform badly in the presence of steep orography: as model resolution increases, steeper gradients lead to larger truncation errors in calculating the horizontal pressure gradientwhich result in spurious winds \citep{dempsey-davis1998} \TODO{or steppeler2002}.

@@

@@

Cut cells is an alternative method in which cells that lie entirely below the terrain are removed, and those that intersect the surface are modified in shape so that they more closely fit the terrain.  This modification means that some cells become very small, which can reduce computational efficiency \TODO{(klein2009)}, and several approaches have been tried to alleviate the problem \TODO{(steppeler2002 thin wall approx,yamazaki-satomura2010 cell combining, jebens2011 implicit method in cut cells)}.

Several studies have found that cut cells produce more accurate results when compared to TF coordinates.  Spurious winds seen in TF coordinates are not present and errors do not increase with steeper terrain \citep{good2014}.  A comparison of TF and cut cells using real initial data by \TODO{steppeler2006} found that precipitation patterns, temperature and wind fields were forecast more accurately in the cut cell model.

This study uses the nonhydrostatic model from \citet{weller-shahrokhi2014} to enable a like-for-like comparison between BTF, SLEVE and cut cell grids for idealised, two-dimensional test cases from the literature.  \TODO{final bit of intro: in section 2, section 3 etc etc}

\begin{itemize}
	\item TF grids often shown to have lesser accuracy than cut cells in some test cases, often significantly so.  Our results demonstrate that accuracy need not be much reduced and can even be better than cut cells. \TODO{this is too confrontational!}
\end{itemize}

@@

\TODO{should we just compare cubicUpwindCPCFit with Sch\"ar's 4th order?  We could also compare our skewCorrected linear with Sch\"ar's linear but I'm not sure what value it would have...}

@@

\begin{itemize}
	\item Motivation: challenge accuracy of H operator (\TODO{if we mention H op here we'll have to explain it beforehand}), curl-free pressure gradient formulation
	\item Cite weller-shahrokhi2014 for domain, thermodynamics, discretisation in time and space
\end{itemize}

Analysis:
\begin{itemize}
	\item Spurious $w$ significantly less than those found by \citet{klemp2011} for BTF: \SI{0.35}{\meter\per\second} vs \SI{10}{\meter\per\second}
	\item SnapCol $w$ (\SI{1e-3}{\meter\per\second}) significantly than BTF or SLEVE
	\item \TODO{Could also compare with zaengl2012, \citet{good2014}?}
	\item Conclusion: non-orthogonality is a significant cause of numerical error in this test but, again, satisfactory on all grids
\end{itemize}

@@

\begin{figure*}
	\centering
	\includegraphics{../msc-publication-fig-gravityWaves/fig-gravityWaves.pdf}
%
	\caption{Gravity wave vertical velocities and thermal anomalies from initial thermal profile}
	\label{fig:gw-w}
\end{figure*}

\begin{figure}
	\includegraphics[height=3in]{gw-theta.png}
	\includegraphics[height=2in]{gw-theta-zoom.png}
%
	\caption{Gravity wave thermal anomalies from initial thermal profile}
	\label{fig:gw-theta}
\end{figure}

@@

\begin{table}[t]
\caption{Min, max and error norms compared with \citet{schaer2002}}
\label{tab:advection}
%
\begin{center}
\begin{tabular}{ l l l l l l }
\hline\hline
& \multicolumn{3}{c}{Cubic upwind-biased} & \multicolumn{2}{c}{Sch\"ar 4th order} \\
& $\ell^2$ error & min & max & min & max \\
\hline
Analytic  & 0 & 0 & 1 & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} \\
BTF 	  & TODO & TODO & TODO & \num{-0.058} & \num{1.001} \\
SnapCol   & TODO & TODO & TODO & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} \\
noOrography & TODO & TODO & TODO & \num{-0.002} & \num{0.984} \\
\hline
\end{tabular}
\end{center}
\end{table}

>>>>

\section{Introduction}
Representing orography accurately in numerical weather prediction systems is necessary to model downslope winds and local precipitation patterns.  There are two main approaches to represent orography on a grid: terrain following layers and cut cells.  Both methods align cells in vertical columns.

With terrain following (TF) layers the terrain's influence decays with height so that the bottommost layers follow the underlying surface closely while the uppermost layers are flat.  TF layers are usually implemented on a rectangular computational grid using transformed coordinates.

It is well-known that TF coordinates perform badly in the presence of steep orography: as model resolution increases, steeper gradients lead to larger truncation errors in calculating the horizontal pressure gradientwhich result in spurious winds \citep{dempsey-davis1998,steppeler2002}.

@@

Cut cells is an alternative method in which cells that lie entirely below the terrain are removed, and those that intersect the surface are modified in shape so that they more closely fit the terrain.  This modification means that some cells become very small, which can reduce computational efficiency \citep{klein2009}, and several approaches have been tried to alleviate the problem \citep{steppeler2002,yamazaki-satomura2010,jebens2011}.

Several studies have found that cut cells produce more accurate results when compared to TF coordinates.  Spurious winds seen in TF coordinates are not present and errors do not increase with steeper terrain \citep{good2014}.  A comparison of TF and cut cells using real initial data by \citet{steppeler2006} found that precipitation patterns, temperature and wind fields were forecast more accurately in the cut cell model.

Many comparisons of TF and cut cell grids have made between different models.  On the contrary, this study uses the nonhydrostatic model from \citet{weller-shahrokhi2014} to enable a like-for-like comparison between BTF, SLEVE and cut cell grids for idealised, two-dimensional test cases from the literature.  \TODO{final bit of intro: in section 2, section 3 etc etc}

\TODO{TF grids often shown to have lesser accuracy than cut cells in some test cases, often significantly so.  Our results demonstrate that accuracy need not be much reduced and can even be better than cut cells.}

\section{Grids}
\citet{galchen-somerville1975} proposed a basic terrain following (BTF) coordinate defined as 
\begin{equation}
	z = \left( H - h \right) \left( z^\star / H \right) + h
\end{equation}
where, in two dimensions, \(z(x, z^\star)\) is the height of the coordinate surface at level \(z^\star\), \(H\) is the height of the domain, and \(h(x)\) is the height of the terrain surface.  Using this coordinate, the terrain's influence decays linearly with height but disappears only at the top of the domain.

@@

\section{Numerical method}

@@

@@

\begin{itemize}
	\item Motivation: challenge curl-free pressure gradient formulation?
	\item Cite weller-shahrokhi2014 for domain, thermodynamics, discretisation in time and space
\end{itemize}

Analysis:
\begin{itemize}
	\item Spurious $w$ significantly less than those found by \citet{klemp2011} for BTF: \SI{0.35}{\meter\per\second} vs \SI{10}{\meter\per\second}
	\item SnapCol $w$ (\SI{1e-3}{\meter\per\second}) significantly than BTF or SLEVE
	\item \TODO{Could also compare with zaengl2012, \citet{good2014}?}
	\item Conclusion: non-orthogonality is a significant cause of numerical error in this test but, again, \TODO{satisfactory?} on all grids
\end{itemize}

@@

\begin{figure}
	\centering
	\includegraphics{../msc-publication-fig-gravityWaves/fig-gravityWaves.pdf}
%
	\caption{Vertical cross section of vertical velocity contours in the gravity waves test after 5 hours on (a) the BTF grid compared with (b) the mass-conserving semi-implicit semi-Lagrangian solution from \citet{melvin2010}.  Contours are every \SI{5e-2}{\meter\per\second} with solid lines denoting ascent and dashed lines descent.}
	\label{fig:gw-w}
\end{figure}

\begin{figure*}
	\centering
	\includegraphics{../msc-publication-fig-gravityWaves-theta/fig-gravityWaves-theta.pdf}
%
	\caption{Anomalies in potential temperature in the gravity waves test after 5 hours.  \TODO{identify subfigures}}
	\label{fig:gw-theta}
\end{figure*}

@@

\begin{table}[t]
\caption{Min, max and error norms compared with \citet{schaer2002}}
\label{tab:advection}
%
\begin{center}
\begin{tabular}{ l l l l l l }
\hline\hline
& \multicolumn{3}{c}{Cubic upwind-biased} & \multicolumn{2}{c}{Sch\"ar 4th order} \\
& $\ell^2$ error & min & max & min & max \\
\hline
Analytic  & 0 & 0 & 1 & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} \\
BTF 	  & TODO & TODO & TODO & \num{-0.058} & \num{1.001} \\
SLEVE 	  & TODO & TODO & TODO & TODO & TODO \\
SnapCol   & TODO & TODO & TODO & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} \\
noOrography & TODO & TODO & TODO & \num{-0.002} & \num{0.984} \\
\hline
\end{tabular}
\end{center}
\end{table}
